---
title: "Bootstrapping"
author: "Makayla Tang"
date: "2022 May"
output: pdf_document
---

```{r setup, include=F}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

In this problem, you will construct an S3 method `bootstrap`, for both the class `numeric` and `stratified` (introduced in Lecture 5), with the following interface.

```{r, include=T, eval=F}
bootstrap.my_class <- function(object, nboot, stat){... your code here ...}
```

The function `bootstrap.my_class` will return the *evaluations* of the statistics (i.e., function) encoded in the function `stat` on each one of the bootstrapped vectors.

Illustrate the use of your `bootstrap` generic function on objects of the class `numeric` and `stratified` using the mean, the median, and the standard deviation as the statistics of interest (e.g. make a histogram with the evaluations of the statistics).

Generalize the methods `bootstrap` defined above to the case of an argument `stat` that is a function that can take additional arguments, e.g. a function that computes the *k*th moment. Test it.

```{r, include=T, eval=T}
bootstrap <- function(object, ...) UseMethod("bootstrap")
```

```{r, include=T, eval=T}
bootstrap.numeric <- function(object, nboot, stat, ...){ 
        if (!is( object, "numeric"))
                stop( "bootstrap.numeric must be 'numeric'" ) 
        if (nboot < 1 | is.infinite(nboot))
                stop( "'nboot' should be a positive integer" ) 
        if (!is( stat, "function"))
                stop( "bootstrap.numeric requires 'stat' of class 'function'" )
        n <- length(object)
        purrr::map_dbl(seq(nboot), function(x) stat(sample(object, size=n, replace=TRUE), ...)) 
}
```

Visualization

```{r, include=T, eval=T}
# Define function moment
moment <- function(x, k) {
        (1/length(x))*sum((x-mean(x))^k)
}

x <- rnorm(100,5,1)
par(mfrow=c(1,4))
hist(bootstrap(x, 1000, mean), main="Mean", xlab="")
hist(bootstrap(x, 100, median), main="Median", xlab="")
hist(bootstrap(x, 100, sd), main="SD", xlab="")
hist(bootstrap(x, 100, moment, k = 2), main="Second moment", xlab="")
```

Constructor function for class *stratified*

```{r, include=T, eval=T}
stratified <- function(y, strata) {
        if (!is.numeric(y)) stop("'y' must be numeric")
        if (!is.factor(strata)) stop("'strata' must be a factor")
        if (length(y) != length(strata)) stop("'y' and 'strata' must have equal length")
        structure(list(y=y, strata=strata), class = "stratified")
}
```

bootstrap method for *stratified* objects

```{r, include=T, eval=T}
bootstrap.stratified <- function(object, nboot, stat, ...){ 
        if ( !is( object, "stratified") )
                stop( "bootstrap.stratified requires an object of class 'stratified'" ) 
        if ( nboot < 1 | is.infinite(nboot) )
                stop( "'nboot' should be a positive integer" ) 
        if ( !is( stat, "function") )
                stop( "bootstrap.numeric requires 'stat' of class 'function'" )
        
        stat_with_args = function(x) stat(x, ...)
        tapply(object$y, object$strata, bootstrap.numeric, nboot, stat_with_args)
}
```

Visualization

```{r, include=T, eval=T}
my_str_samp <- stratified(y = c(rgamma(50,3), rnorm(100, 30)),
                          strata = factor(c(rep("a",50),rep("b",100))))
bootstrap(my_str_samp, 10, mean)
bootstrap(my_str_samp, 10, median)
bootstrap(my_str_samp, 10, sd)
bootstrap(my_str_samp, 10, moment, 2)
```

